\chapter{Python Documentation}

\section{Purpose of the Documentation}

The project includes a separate Python documentation site built with Sphinx.
The purpose of this documentation is to make the implementation inspectable at module and API level, without overloading the main report with low-level interface details.
The report and the Sphinx documentation therefore have complementary roles:

\begin{itemize}
    \item the report explains the mathematical model, experimental design, software architecture, and empirical results;
    \item the Sphinx documentation exposes the package structure, module-level documentation, function signatures, class interfaces, and docstrings.
\end{itemize}
This separation keeps the report focused on the scientific content while still making the codebase auditable.

\section{Sphinx Documentation Structure}

The documentation sources are stored in:

\begin{projectlistingbox}
docs/source/
\end{projectlistingbox}
\projectlistingcaption{Source directory for the Sphinx documentation.}

The API reference is generated from the docstrings contained in the Python package:

\begin{projectlistingbox}
src/nn_option_pricing/
\end{projectlistingbox}
\projectlistingcaption{Main Python package documented through Sphinx autodoc.}

The most important documentation entry points are summarized in Table~\ref{tab:appendix-sphinx-entry-points}.

\begin{table}[H]
    \centering
    \small
    \begin{tabular}{@{}
        >{\raggedright\arraybackslash}p{0.33\linewidth}
        >{\raggedright\arraybackslash}p{0.52\linewidth}
        @{}}
        \toprule
        File or directory                          & Role                                           \\
        \midrule
        \path{docs/source/index.rst}               & main documentation index                       \\
        \path{docs/source/overview.rst}            & project orientation and repository overview    \\
        \path{docs/source/methodology.rst}         & methodological description of the experiment   \\
        \path{docs/source/experiment_pipeline.rst} & end-to-end experiment pipeline                 \\
        \path{docs/source/experiments.rst}         & additional experiment descriptions             \\
        \path{docs/source/testing.rst}             & testing and verification strategy              \\
        \path{docs/source/api/}                    & generated API reference for the Python package \\
        \bottomrule
    \end{tabular}
    \caption{Main Sphinx documentation entry points.}
    \label{tab:appendix-sphinx-entry-points}
\end{table}

\section{Local Documentation Build}

The documentation can be built locally from the repository root after installing the project and documentation dependencies:

\begin{projectlistingbox}[shellstyle]
pip install -r requirements.txt
pip install -r requirements-docs.txt
pip install -e . --no-build-isolation
\end{projectlistingbox}
\projectlistingcaption{Installation commands for local documentation builds.}

The standard HTML build command is:

\begin{projectlistingbox}[shellstyle]
sphinx-build -b html docs/source docs/build/html
\end{projectlistingbox}
\projectlistingcaption{Local Sphinx HTML build command.}

The generated documentation can then be opened from:

\begin{projectlistingbox}
docs/build/html/index.html
\end{projectlistingbox}
\projectlistingcaption{Local HTML documentation entry point.}

\section{Strict Documentation Build}

During development, it is useful to treat documentation warnings as errors.
This makes missing references, malformed directives, or broken API documentation visible immediately:

\begin{projectlistingbox}[shellstyle]
sphinx-build -W -b html docs/source docs/build/html
\end{projectlistingbox}
\projectlistingcaption{Strict Sphinx build command.}

The same strict build is used in the automatic documentation workflow described below.

\section{Automatic Deployment with GitHub Actions and GitHub Pages}

The generated HTML documentation is not tracked in the repository.
This is intentional: \path{docs/build/html/} is a generated artifact, and keeping it out of version control avoids unnecessary file churn.
Instead, the documentation can be built automatically through GitHub Actions and published with GitHub Pages.

The workflow is stored in:

\begin{projectlistingbox}
.github/workflows/docs.yml
\end{projectlistingbox}
\projectlistingcaption{GitHub Actions workflow for building and publishing the documentation.}

At each push to the main branch, the workflow performs the following operations:

\begin{enumerate}
    \item checks out the repository;
    \item installs Python and the project dependencies;
    \item installs the documentation dependencies;
    \item builds the Sphinx documentation using the strict \texttt{-W} option;
    \item uploads the generated HTML site as a GitHub Pages artifact;
    \item deploys the artifact to GitHub Pages.
\end{enumerate}
This approach keeps the repository clean while still providing an online version of the documentation.
Once GitHub Pages is enabled for the repository, the documentation is available through the repository Pages URL.

The main online resources are reported in Table~\ref{tab:appendix-online-resources}.

\begin{table}[H]
    \centering
    \small
    \begin{tabular}{@{}
        >{\raggedright\arraybackslash}p{0.28\linewidth}
        >{\raggedright\arraybackslash}p{0.57\linewidth}
        @{}}
        \toprule
        Resource             & URL                                                     \\
        \midrule
        GitHub repository    & \url{https://github.com/matteogiorgi/nn-option-pricing} \\
        Online documentation & \url{https://matteogiorgi.github.io/nn-option-pricing}  \\
        \bottomrule
    \end{tabular}
    \caption{Online repository and documentation resources.}
    \label{tab:appendix-online-resources}
\end{table}
