\chapter{Code Navigation}

\section{How to Read the Codebase}

The implementation is organized as a reusable Python package plus a small set of command-line scripts.
For a reader who wants to inspect the code without reading every file line by line, the most useful path is:

\begin{enumerate}
    \item start from \href{https://github.com/matteogiorgi/nn-option-pricing/blob/main/scripts/run_experiment.py}{\path{scripts/run_experiment.py}}, which defines the command-line interface for the main experiment and calls the main pipeline;
    \item inspect \href{https://github.com/matteogiorgi/nn-option-pricing/blob/main/src/nn_option_pricing/config.py}{\path{src/nn_option_pricing/config.py}}, where experiment parameters are represented as immutable configuration objects;
    \item follow \href{https://github.com/matteogiorgi/nn-option-pricing/blob/main/src/nn_option_pricing/pipeline.py}{\path{src/nn_option_pricing/pipeline.py}}, which connects the main experimental stages and saves the results;
    \item inspect the specialized modules only when needed.
\end{enumerate}
This gives a top-down view of the project before moving into the individual numerical and machine-learning components.

\section{Main Code Entry Points}

The following list summarizes the most important files for understanding the implementation.
Except for \path{run_experiment.py}, which is in \path{scripts/}, the listed modules are located in \path{src/nn_option_pricing/}.
The file names are clickable links to the corresponding GitHub source files.
For stable archival references, the \texttt{main} branch in these URLs can be replaced by the final commit hash.

\begin{itemize}
    \item \href{https://github.com/matteogiorgi/nn-option-pricing/blob/main/scripts/run_experiment.py}{\path{run_experiment.py}}: main experiment command-line interface.
    \item \href{https://github.com/matteogiorgi/nn-option-pricing/blob/main/src/nn_option_pricing/config.py}{\path{config.py}}: experiment configuration dataclasses and output paths.
    \item \href{https://github.com/matteogiorgi/nn-option-pricing/blob/main/src/nn_option_pricing/pipeline.py}{\path{pipeline.py}}: end-to-end clean-target neural-network experiment.
    \item \href{https://github.com/matteogiorgi/nn-option-pricing/blob/main/src/nn_option_pricing/model.py}{\path{model.py}}: feed-forward neural-network architecture.
    \item \href{https://github.com/matteogiorgi/nn-option-pricing/blob/main/src/nn_option_pricing/black_scholes.py}{\path{black_scholes.py}}: analytical Black-Scholes pricing and payoff functions.
    \item \href{https://github.com/matteogiorgi/nn-option-pricing/blob/main/src/nn_option_pricing/dataset.py}{\path{dataset.py}}: synthetic dataset generation and feature construction.
    \item \href{https://github.com/matteogiorgi/nn-option-pricing/blob/main/src/nn_option_pricing/training.py}{\path{training.py}}: neural-network training, scaling, prediction, and early stopping.
    \item \href{https://github.com/matteogiorgi/nn-option-pricing/blob/main/src/nn_option_pricing/monte_carlo.py}{\path{monte_carlo.py}}: batched Monte Carlo benchmark.
    \item \href{https://github.com/matteogiorgi/nn-option-pricing/blob/main/src/nn_option_pricing/evaluation.py}{\path{evaluation.py}}: regression metrics used in the experiments.
    \item \href{https://github.com/matteogiorgi/nn-option-pricing/blob/main/src/nn_option_pricing/plots.py}{\path{plots.py}}: diagnostic plots saved by the pipeline.
    \item \href{https://github.com/matteogiorgi/nn-option-pricing/blob/main/src/nn_option_pricing/svr.py}{\path{svr.py}}: reduced-scale Support Vector Regression benchmark.
    \item \href{https://github.com/matteogiorgi/nn-option-pricing/blob/main/src/nn_option_pricing/noise.py}{\path{noise.py}}: controlled target-noise generation.
    \item \href{https://github.com/matteogiorgi/nn-option-pricing/blob/main/src/nn_option_pricing/noisy_experiment.py}{\path{noisy_experiment.py}}: noisy-target neural-network experiment.
    \item \href{https://github.com/matteogiorgi/nn-option-pricing/blob/main/src/nn_option_pricing/noisy_svr_experiment.py}{\path{noisy_svr_experiment.py}}: noisy-target SVR benchmark.
\end{itemize}

\section{Experiment Scripts}

The repository contains separate scripts for the main experiment and for additional benchmarks.
This keeps the main pipeline focused while making each extension reproducible.

\begin{itemize}
    \item \href{https://github.com/matteogiorgi/nn-option-pricing/blob/main/scripts/run_experiment.py}{\path{run_experiment.py}}: main neural-network experiment.
    \item \href{https://github.com/matteogiorgi/nn-option-pricing/blob/main/scripts/benchmark_runtime.py}{\path{benchmark_runtime.py}}: pricing runtime comparison.
    \item \href{https://github.com/matteogiorgi/nn-option-pricing/blob/main/scripts/run_svr_benchmark.py}{\path{run_svr_benchmark.py}}: clean-target SVR benchmark.
    \item \href{https://github.com/matteogiorgi/nn-option-pricing/blob/main/scripts/run_noisy_targets_experiment.py}{\path{run_noisy_targets_experiment.py}}: noisy-target neural-network experiment.
    \item \href{https://github.com/matteogiorgi/nn-option-pricing/blob/main/scripts/run_noisy_svr_benchmark.py}{\path{run_noisy_svr_benchmark.py}}: noisy-target SVR benchmark.
\end{itemize}
