\chapter{Financial and Mathematical Background}

\section{European Call Options}

A European call option gives its holder the right, but not the obligation, to buy an underlying asset at a fixed strike price $K$ at maturity $T$.
The term European means that the option can be exercised only at maturity, not before.
If the terminal asset price is $S_T$, the payoff of a European call is:

\begin{equation}
    \Phi(S_T) = \max(S_T - K, 0) = (S_T - K)^+.
\end{equation}

The payoff is positive only when the option finishes in the money, namely when $S_T > K$.
If $S_T \leq K$, exercising the option would not be convenient and the payoff is zero.
The option value depends on the current asset price, the strike price, time to maturity, the risk-free rate, and the volatility of the underlying asset.
General arbitrage-pricing principles for derivative securities can be introduced in discrete-time models before moving to continuous-time diffusion models \cite{shreve2004stochastic1,pascucci2012financial}.
In this project, only European call options are considered.

Before introducing the stochastic model, Table~\ref{tab:financial-notation} fixes the notation that will be used in the rest of the report.

\begin{table}[H]
    \centering
    \begin{tabular}{ll}
        \toprule
        Symbol & Meaning \\
        \midrule
        $S_0$ & initial price of the underlying asset \\
        $S_t$ & price of the underlying asset at time $t$ \\
        $K$ & strike price of the option \\
        $T$ & time to maturity, expressed in years \\
        $r$ & continuously compounded risk-free interest rate \\
        $\sigma$ & volatility of the underlying asset \\
        $W_t$ & standard Brownian motion \\
        $\mathbb{Q}$ & risk-neutral probability measure \\
        $\Phi(S_T)$ & payoff of the option at maturity \\
        $C_{BS}$ & analytical Black-Scholes call price \\
        $\hat{C}_{MC}$ & Monte Carlo estimate of the call price \\
        \bottomrule
    \end{tabular}
    \caption{Main financial notation used in the report.}
    \label{tab:financial-notation}
\end{table}

\section{Stochastic Differential Equations}

Continuous-time asset price models are often described by stochastic differential equations.
An SDE combines a deterministic drift component with a random diffusion component driven by Brownian motion.
This framework is standard in financial modeling because asset prices evolve continuously but are subject to uncertainty \cite{shreve2004stochastic}.
In general form, an Itô diffusion can be written as:

\begin{equation}
    dX_t = \mu(t, X_t)\,dt + \eta(t, X_t)\,dW_t,
\end{equation}

where $\mu$ is the drift coefficient, $\eta$ is the diffusion coefficient, and $W_t$ is a Brownian motion.
The term $\mu(t, X_t)\,dt$ describes the deterministic infinitesimal trend of the process, while $\eta(t, X_t)\,dW_t$ describes the random infinitesimal shock.
In financial models, this decomposition is useful because asset prices display both systematic growth components and random fluctuations.

\section{Geometric Brownian Motion}

In the Black-Scholes model, the underlying asset follows a geometric Brownian motion.
Under the risk-neutral measure $\mathbb{Q}$, the discounted asset price is a martingale; equivalently, the asset drift in the pricing dynamics is the risk-free rate $r$.
The dynamics used for pricing are:

\begin{equation}
    dS_t = r S_t dt + \sigma S_t dW_t,
\end{equation}

where $S_t$ is the asset price at time $t$, $r$ is the risk-free interest rate, $\sigma$ is the constant volatility, and $W_t$ is a standard Brownian motion.
The multiplicative structure of the SDE means that both the drift and the volatility scale with the current asset price.
As a consequence, if $S_0 > 0$, the asset price remains positive under the model.

The solution at maturity $T$ is:

\begin{equation}
    S_T = S_0 \exp\left[\left(r - \frac{1}{2}\sigma^2\right)T + \sigma \sqrt{T} Z\right],
    \qquad Z \sim \mathcal{N}(0,1).
\end{equation}

This explicit terminal distribution is important for the implementation.
It allows Monte Carlo simulation to sample $S_T$ directly, without simulating all intermediate time steps of the path.

\section{Black-Scholes Model}

The Black-Scholes model assumes frictionless markets, constant volatility, a constant risk-free rate, no arbitrage, and continuous trading \cite{black1973pricing,shreve2004stochastic}.
Under these assumptions, the price of a European option can be derived from risk-neutral valuation.
The main idea is that, in an arbitrage-free complete market, the option price is equal to the discounted expected payoff under the risk-neutral measure.

For a European call, the value is the discounted expected payoff:

\begin{equation}
    C = e^{-rT}\mathbb{E}^{\mathbb{Q}}\left[\max(S_T - K, 0)\right].
\end{equation}

\begin{quote}
    In this project, Black-Scholes is not treated as an empirical description of all market prices.
    It is used as a controlled mathematical pricing function that provides exact target values for supervised learning.
\end{quote}

\section{Black-Scholes Formula}

The analytical Black-Scholes price of a European call option is:

\begin{equation}
    C_{BS} = S_0 N(d_1) - K e^{-rT} N(d_2),
\end{equation}

where $N(\cdot)$ is the cumulative distribution function of a standard normal random variable and:

\begin{align}
    d_1 & = \frac{\ln(S_0/K) + \left(r + \frac{1}{2}\sigma^2\right)T}{\sigma\sqrt{T}}, \\
    d_2 & = d_1 - \sigma\sqrt{T}.
\end{align}

In this project, $C_{BS}$ is used both as the supervised learning target and as the reference value for evaluation.
This is what makes the experiment controlled: the true target function is known analytically.
The neural network is therefore trained to approximate the mapping:

\begin{equation}
    f_{BS}: (S_0, K, T, r, \sigma) \mapsto C_{BS}.
\end{equation}

This formulation turns option pricing into a supervised regression problem.
Each synthetic observation contains a vector of financial parameters as input and the corresponding analytical Black-Scholes price as target.

\section{Monte Carlo Pricing}

Monte Carlo simulation approximates the risk-neutral expectation by generating many realizations of $S_T$ \cite{glasserman2004monte}:

\begin{equation}
    \hat{C}_{MC} =
    e^{-rT}
    \frac{1}{M}
    \sum_{i=1}^{M}
    \max(S_T^{(i)} - K, 0).
\end{equation}

For sufficiently large $M$, $\hat{C}_{MC}$ converges to $C_{BS}$ by the law of large numbers.
The statistical error decreases at the standard Monte Carlo rate:

\begin{equation}
    \operatorname{SE}(\hat{C}_{MC}) =
    \frac{e^{-rT}}{\sqrt{M}}
    \operatorname{Std}\left(\max(S_T - K, 0)\right).
\end{equation}

This $M^{-1/2}$ convergence rate explains why Monte Carlo can become expensive when high accuracy is required.
In this project, Monte Carlo is used as a numerical benchmark and as a comparison point for the neural network.
Since the terminal distribution of $S_T$ is known exactly under Black-Scholes, the implementation simulates terminal prices directly rather than discretizing the full path.
This choice is both mathematically correct under the Black-Scholes assumptions and computationally more efficient than Euler discretization for the same model.
