\chapter{Experiments}

\section{Final Experiment Design}

The final experiment was designed to evaluate the neural network as a supervised surrogate of the Black-Scholes European call pricing function on a sufficiently large synthetic dataset.
The analytical Black-Scholes formula is used as the reference target, while Monte Carlo simulation is used as a numerical benchmark on a deterministic subset of the test set.

The final run used 100,000 synthetic option contracts sampled from the parameter ranges defined in the methodology.
The selected neural network configuration includes moneyness $S_0/K$ as an engineered input feature and uses the SiLU activation function in the hidden layers.
The neural network was trained for up to 200 epochs with early stopping, using mini-batches of size 1024.
The Monte Carlo benchmark used 50,000 simulated paths on 512 test options.

\begin{lstlisting}[style=pythonstyle, caption={Command used for the final experiment.}]
.venv/bin/python scripts/run_experiment.py \
  --n-samples 100000 \
  --max-epochs 200 \
  --batch-size 1024 \
  --mc-n-paths 50000 \
  --mc-evaluation-samples 512 \
  --feature-set with_moneyness \
  --activation silu \
  --seed 42 \
  --data-dir data/final_improved \
  --output-dir outputs/final_improved
\end{lstlisting}

\section{Reproducible Artifacts}

The complete generated dataset, trained model checkpoint, and fitted scalers are stored under \texttt{data/final\_improved/} and \texttt{outputs/final\_improved/}.
These directories are ignored by Git because they can be reproduced from the code and experiment configuration.

The selected final artifacts used in the report are tracked in \texttt{results/final/}.
This directory contains:

\begin{itemize}
    \item the experiment configuration snapshot;
    \item neural network metrics;
    \item Monte Carlo benchmark metrics;
    \item diagnostic figures;
    \item a short verification note reporting the final test and documentation checks.
\end{itemize}

\section{Evaluation Protocol}

The neural network is evaluated on a held-out test set that is not used during training or validation.
The target prices are analytical Black-Scholes prices, so the reported errors measure how accurately the network approximates the known pricing function.

The Monte Carlo benchmark is evaluated against the same analytical Black-Scholes reference, but only on a deterministic subset of 512 test options.
This restriction keeps the numerical benchmark computationally bounded while still providing a useful comparison with a classical simulation-based method.

The main metrics are mean absolute error, root mean squared error, coefficient of determination, and MAPE computed only on prices greater than 1.
The MAPE restriction avoids unstable percentage errors for options whose theoretical price is close to zero.

\section{Support Vector Regression Baseline}

As an additional classical machine learning baseline, the project includes a reduced-scale Support Vector Regression experiment with an RBF kernel.
This benchmark is not intended to replace the main neural network experiment.
Instead, it provides a useful comparison with a non-neural supervised regression method.

The SVR benchmark is evaluated on smaller synthetic datasets because kernel methods scale less favorably than neural networks as the number of training observations increases.
For this reason, the experiment uses 5,000 synthetic samples per run and is repeated over three random seeds.
The same engineered feature set used in the final neural network experiment, including moneyness $S_0/K$, is used for the SVR baseline.
Both input features and target prices are standardized before fitting, and predictions are transformed back to original price units before evaluation.

\begin{lstlisting}[style=pythonstyle, caption={Command used for the SVR benchmark.}]
.venv/bin/python scripts/run_svr_benchmark.py \
  --n-samples 5000 \
  --seeds 11 42 73 \
  --feature-set with_moneyness \
  --output-dir results/experiments/svr_benchmark
\end{lstlisting}

\section{Diagnostic Visualizations}

The experiment produces the following diagnostic plots:

\begin{itemize}
    \item training loss and validation loss;
    \item comparison between the Black-Scholes price and the neural network prediction;
    \item distribution of prediction errors;
    \item absolute error as a function of moneyness $S_0/K$;
    \item absolute error as a function of maturity $T$;
    \item absolute error as a function of volatility $\sigma$.
\end{itemize}
