\chapter{Software Implementation}

\section{Implementation Goals}

The project is implemented as a small Python package rather than as a single script.
This design choice keeps the mathematical components, training logic, benchmarks, plotting functions, and command-line interfaces separated.
The main implementation goals are:

\begin{itemize}
    \item reproducibility: experiments are controlled by explicit configuration objects and random seeds;
    \item modularity: each file has a focused responsibility;
    \item efficiency: vectorized NumPy operations, batched Monte Carlo simulation, and mini-batch neural-network training are used;
    \item inspectability: metrics, figures, datasets, models, scalers, and configuration snapshots are saved to disk.
\end{itemize}
The main libraries are NumPy, pandas, SciPy, scikit-learn, matplotlib, and PyTorch.

\begin{quote}
    The codebase is organized around a reusable scientific package.
    Command-line scripts are thin entry points that configure and execute package functions.
\end{quote}

\section{Repository Structure}

The reusable scientific and machine-learning logic is contained in \path{src/nn_option_pricing/}.
The files in \texttt{scripts/} are command-line entry points that build configurations and call functions from the package.

\begin{projectlistingbox}[shellstyle]
src/nn_option_pricing/
    black_scholes.py          # analytical formula and payoff
    config.py                 # experiment configuration dataclasses
    dataset.py                # synthetic data generation and storage
    evaluation.py             # regression metrics
    model.py                  # PyTorch feed-forward neural network
    monte_carlo.py            # efficient Monte Carlo simulation
    noise.py                  # controlled target-noise generation
    noisy_experiment.py       # neural-network noisy-target experiment
    noisy_svr_experiment.py   # noisy-target SVR benchmark
    pipeline.py               # end-to-end experiment
    plots.py                  # visualizations
    svr.py                    # reduced-scale SVR benchmark
    training.py               # training loop, scaling, early stopping
scripts/
    run_experiment.py             # main experiment CLI
    benchmark_runtime.py          # runtime benchmark CLI
    run_svr_benchmark.py          # SVR benchmark CLI
    run_noisy_targets_experiment.py
    run_noisy_svr_benchmark.py
tests/
    test_black_scholes.py
    test_dataset.py
    test_model.py
    test_monte_carlo.py
    test_noise.py
    test_noisy_svr.py
    test_pipeline.py
    test_svr.py
\end{projectlistingbox}
\projectlistingcaption{Main source, script, and test structure.}

Table~\ref{tab:software-modules} summarizes the role of the most important implementation modules.

\begin{table}[H]
    \centering
    \begin{tabular}{lll}
        \toprule
        Module & Main responsibility & Used by \\
        \midrule
        \texttt{black\_scholes.py} & analytical pricing and payoff & dataset, benchmarks \\
        \texttt{dataset.py} & synthetic data generation & pipeline, SVR experiments \\
        \texttt{training.py} & scaling, training, prediction & main and noisy NN experiments \\
        \texttt{monte\_carlo.py} & batched Monte Carlo estimator & pipeline, runtime benchmark \\
        \texttt{evaluation.py} & regression metrics & all experiments \\
        \texttt{pipeline.py} & end-to-end main run & \texttt{run\_experiment.py} \\
        \texttt{svr.py} & reduced-scale SVR benchmark & \texttt{run\_svr\_benchmark.py} \\
        \texttt{noise.py} & controlled label noise & noisy-target experiments \\
        \bottomrule
    \end{tabular}
    \caption{Main Python modules and their responsibilities.}
    \label{tab:software-modules}
\end{table}

\section{Configuration System}

The project uses frozen dataclasses to keep experiment parameters explicit.
The configuration layer includes dataset ranges, training hyperparameters, Monte Carlo settings, and output paths.
This avoids hard-coded constants scattered across the codebase and makes experiments easier to reproduce.

The top-level configuration groups the main components:

\begin{projectlistingbox}
@dataclass(frozen=True)
class ExperimentConfig:
    dataset: DatasetConfig = DatasetConfig()
    training: TrainingConfig = TrainingConfig()
    monte_carlo: MonteCarloConfig = MonteCarloConfig()
    paths: PathConfig = PathConfig()
\end{projectlistingbox}
\projectlistingcaption{Top-level experiment configuration object.}

Each experiment saves a configuration snapshot in JSON format.
Since dataclasses and \texttt{Path} objects are not directly JSON-serializable, the helper \texttt{config\_to\_dict} recursively converts configuration objects into JSON-compatible values.
This makes it possible to recover the exact settings used to produce a result.

\section{Command-Line Interfaces}

The project is executed through command-line scripts.
The main experiment is launched by the script \texttt{run\_experiment.py}.
It parses command-line arguments, builds an \texttt{ExperimentConfig}, runs the pipeline, and prints the resulting metrics.

\begin{projectlistingbox}
config = ExperimentConfig(
    dataset=DatasetConfig(n_samples=args.n_samples, seed=args.seed),
    training=TrainingConfig(
        seed=args.seed,
        feature_set=args.feature_set,
        max_epochs=args.max_epochs,
        batch_size=args.batch_size,
        learning_rate=args.learning_rate,
        activation=args.activation,
    ),
    monte_carlo=MonteCarloConfig(
        n_paths=args.mc_n_paths,
        evaluation_samples=args.mc_evaluation_samples,
    ),
    paths=make_path_config(data_dir=args.data_dir, output_dir=args.output_dir),
)
\end{projectlistingbox}
\projectlistingcaption{Construction of the main experiment configuration from CLI arguments.}

The final experiment is run through:

\begin{projectlistingbox}[shellstyle]
.venv/bin/python scripts/run_experiment.py \
    --n-samples 100000 \
    --max-epochs 200 \
    --batch-size 1024 \
    --mc-n-paths 50000 \
    --mc-evaluation-samples 512 \
    --feature-set with_moneyness \
    --activation silu \
    --seed 42 \
    --data-dir data/final_improved \
    --output-dir outputs/final_improved
\end{projectlistingbox}
\projectlistingcaption{Command used for the final experiment.}

Additional scripts run the runtime benchmark, the SVR benchmark, the noisy-target neural-network experiment, and the noisy-target SVR benchmark.
This keeps the main experiment clean while still allowing extensions to be reproduced independently.

\section{End-to-End Pipeline}

The main pipeline connects all core components.
It is implemented in \texttt{pipeline.py} and performs the following steps:

\begin{enumerate}
    \item create output directories;
    \item save the experiment configuration;
    \item generate and save the synthetic dataset;
    \item train the neural network;
    \item predict on the held-out test set;
    \item compute neural-network metrics;
    \item save the trained model, input scaler, and target scaler;
    \item run the Monte Carlo benchmark on a deterministic test subset;
    \item save metrics and diagnostic plots.
\end{enumerate}
This structure ensures that each run leaves behind enough information to inspect, reproduce, and discuss the experiment.

Table~\ref{tab:software-artifacts} reports the main artifacts produced by the pipeline.

\begin{table}[H]
    \centering
    \begin{tabular}{lll}
        \toprule
        Artifact & Typical location & Purpose \\
        \midrule
        Dataset CSV & \texttt{data/.../synthetic\_options.csv} & generated observations \\
        Configuration JSON & \texttt{outputs/.../experiment\_config.json} & reproducibility \\
        Model checkpoint & \texttt{outputs/.../model.pt} & trained network state \\
        Input scaler & \texttt{outputs/.../scaler.joblib} & feature preprocessing \\
        Target scaler & \texttt{outputs/.../target\_scaler.joblib} & inverse price scaling \\
        Metrics JSON & \texttt{outputs/.../metrics/} & quantitative evaluation \\
        Figures & \texttt{outputs/.../figures/} & diagnostic plots \\
        \bottomrule
    \end{tabular}
    \caption{Main artifacts produced by a complete experiment run.}
    \label{tab:software-artifacts}
\end{table}

\section{Testing and Documentation}

The test suite checks both low-level mathematical functions and higher-level experiment behavior.
This is important because numerical projects can fail silently if only final metrics are inspected.

\begin{table}[H]
    \centering
    \begin{tabular}{ll}
        \toprule
        Test area & Purpose \\
        \midrule
        Black-Scholes tests & validate analytical pricing behavior \\
        Dataset tests & check generated columns, ranges, and shape \\
        Model tests & verify neural-network construction and outputs \\
        Monte Carlo tests & check simulation behavior and convergence properties \\
        Noise tests & validate controlled noisy-target generation \\
        SVR tests & check reduced-scale benchmark execution \\
        Noisy-SVR tests & check noisy-label SVR benchmark execution \\
        Pipeline tests & run small end-to-end smoke tests \\
        \bottomrule
    \end{tabular}
    \caption{Main test areas covered by the project.}
    \label{tab:test-coverage}
\end{table}

The project also includes Sphinx documentation generated from narrative documentation and Python docstrings.
This supports code readability and makes the implementation easier to inspect independently from the report.

\section{Computational Efficiency}

Several implementation choices are made for computational efficiency.
The Black-Scholes formula and dataset generation are vectorized with NumPy.
This allows prices for many contracts to be computed at once without explicit Python loops.

The Monte Carlo method uses the exact terminal distribution of geometric Brownian motion, avoiding unnecessary time discretization.
Simulation is batched both over options and over paths, which keeps memory usage bounded.
If $B_o$ is the option batch size and $B_p$ is the path batch size, the largest simulated terminal-price block has shape approximately:

\begin{equation}
    B_o \times B_p.
\end{equation}

The neural network training uses PyTorch \texttt{DataLoader}, mini-batch gradient descent, input standardization, and target scaling.
Prediction is performed in batches under \texttt{torch.no\_grad()}, so gradients are not stored during inference.
These choices make the code suitable both for quick development runs and for larger experiments.
