\chapter{Conclusions and Next Steps}

The final experiment confirms that a feed-forward neural network can approximate the Black-Scholes pricing function for European call options with high accuracy in a controlled synthetic environment.
With 100,000 synthetic contracts, the network achieves an MAE of 0.04290, an RMSE of 0.06208, and an $R^2$ of 0.999993 on the held-out test set.
These results answer the research question positively within the assumptions and parameter ranges considered in the project.

The comparison with Monte Carlo also clarifies the role of the neural network.
Monte Carlo is a classical numerical estimator and remains useful as a benchmark, while the trained neural network acts as a deterministic pricing surrogate.
The purpose of the network is therefore not to replace the analytical Black-Scholes formula, but to demonstrate that a supervised model can accurately learn a well-defined pricing map.
The runtime benchmark further supports the surrogate interpretation: after training, neural inference is substantially faster than Monte Carlo simulation for repeated pricing queries.
The additional SVR experiment provides a classical machine learning reference point on reduced datasets and supports the choice of the neural network as the main scalable surrogate model.
The noisy-target experiment further shows that the network remains accurate under controlled label perturbations, although approximation quality decreases as the noise level increases.

The main limitations are:

\begin{itemize}
    \item the data are synthetic and generated under Black-Scholes assumptions;
    \item only European call options are considered;
    \item the target pricing function is known analytically;
    \item noisy targets are controlled perturbations and should not be interpreted as real market prices;
    \item generalization outside the sampled parameter ranges is not guaranteed.
\end{itemize}

Natural next steps are:

\begin{itemize}
    \item compare additional neural network architectures beyond the selected MLP;
    \item tune and evaluate further classical machine learning baselines on controlled reduced datasets;
    \item study the error in specific regions of the parameter space, for instance deep in-the-money or out-of-the-money options;
    \item extend the framework to stochastic volatility models such as Heston;
    \item estimate Greeks using either automatic differentiation or finite differences.
\end{itemize}

These extensions are outside the scope of the current implementation, but they represent natural directions for generalizing the approach beyond the Black-Scholes benchmark.
