\chapter{Discussion}

\section{Answer to the Research Question}

The research question of the project is whether a feed-forward neural network can approximate the Black-Scholes pricing function for European call options.
The experimental evidence supports a positive answer within the sampled parameter domain and under the assumptions of the Black-Scholes model.

In mathematical terms, the project studies the approximation problem

\begin{equation}
    f_\theta(x)
    \approx
    C_{BS}(x),
    \qquad
    x = (S_0, K, T, r, \sigma),
\end{equation}

or, in the final configuration,

\begin{equation}
    f_\theta(S_0, K, T, r, \sigma, S_0/K)
    \approx
    C_{BS}(S_0, K, T, r, \sigma).
\end{equation}

The final test-set results show very low absolute error and an $R^2$ close to one.
This indicates that, on held-out synthetic contracts sampled from the same distribution as the training data, the trained neural network reproduces the analytical pricing map very accurately.

\begin{quote}
    The strongest conclusion of the project is conditional but meaningful:
    within the Black-Scholes synthetic setting, the neural network is an accurate learned approximation of the analytical pricing function.
\end{quote}

This distinction is important.
The project does not show that a neural network can price arbitrary real market options.
It shows that, when the target pricing rule is known and deterministic, a supervised neural network can learn that rule with high accuracy and then evaluate it quickly.

\section{What the Results Support}

The results support three main claims, but each claim has to be read together with its evidence and its caveat.
This interpretation is organized in Table~\ref{tab:discussion-claims}.

\begin{table}[H]
    \centering
    \begin{tabular}{
        >{\raggedright\arraybackslash}p{0.27\linewidth}
        >{\raggedright\arraybackslash}p{0.37\linewidth}
        >{\raggedright\arraybackslash}p{0.26\linewidth}
        }
        \toprule
        Claim & Evidence                                                                               & Caveat \\
        \midrule
        The neural network approximates Black-Scholes accurately
              & final MAE, RMSE, $R^2$, true-vs-predicted plot, and error diagnostics
              & valid on the sampled parameter domain, not automatically outside it                             \\
        \projecttablerowsep
        Neural inference is useful as a surrogate
              & runtime benchmark shows much faster inference than Monte Carlo
              & training has an upfront cost and the analytical formula is still faster when available          \\
        \projecttablerowsep
        Controlled noise degrades performance predictably
              & noisy-target experiments show increasing MAE and RMSE with increasing label noise
              & the noise is synthetic and should not be interpreted as real market microstructure              \\
        \bottomrule
    \end{tabular}
    \caption{Interpretation of the main empirical claims.}
    \label{tab:discussion-claims}
\end{table}

The first claim is the central one.
The final model produces predictions that are close to the analytical Black-Scholes prices on the test set.
This confirms that a relatively small MLP can approximate a smooth financial pricing function when the training data cover the relevant parameter ranges.

The second claim concerns computational role rather than mathematical exactness.
The neural network is not proposed as a replacement for the closed-form Black-Scholes formula.
When a closed-form formula exists, it remains exact and extremely fast.
The more relevant comparison is with simulation-based pricing methods such as Monte Carlo, where repeated evaluation can be more expensive.

The third claim concerns robustness.
Adding controlled label noise makes the learning problem harder, but the neural network does not collapse under moderate noise.
This suggests that the model can partially smooth noisy supervision, although the experiment remains deliberately synthetic and controlled.

\section{Role of Monte Carlo}

Monte Carlo simulation plays two roles in the project.
First, it is a classical quantitative benchmark for option pricing.
Second, it provides a natural comparison for the surrogate-model idea.

For a European call option under Black-Scholes dynamics, Monte Carlo estimates

\begin{equation}
    C_{BS}
    =
    e^{-rT}
    \mathbb{E}^{\mathbb{Q}}\left[(S_T-K)^+\right]
\end{equation}

by the sample average

\begin{equation}
    \hat{C}_{MC}
    =
    e^{-rT}
    \frac{1}{M}
    \sum_{j=1}^{M}
    (S_T^{(j)}-K)^+.
\end{equation}

Its error decreases at the standard rate $O_P(M^{-1/2})$.
This convergence rate is robust and model-independent, but it also explains why Monte Carlo can be computationally expensive when many paths or many contracts are required.

The neural network follows a different logic.
It pays a training cost upfront, then produces prices through a deterministic forward pass:

\begin{equation}
    \hat{C}_{NN} = f_\theta(x).
\end{equation}

Therefore, the runtime comparison should be read as a comparison between repeated simulation and learned inference after training.
It should not be read as evidence that the neural network is better than Black-Scholes itself, because the analytical formula remains the natural method when it is available.

\section{Role of Feature Engineering and Activation Choice}

The intermediate experiments help explain why the final configuration was selected.
Adding moneyness exposes a financially meaningful ratio:

\begin{equation}
    m = \frac{S_0}{K}.
\end{equation}

This ratio does not add new information, since it is computed from $S_0$ and $K$.
However, it gives the neural network direct access to a quantity that strongly affects the payoff regime of a call option.
In practical terms, moneyness helps distinguish out-of-the-money, at-the-money, and in-the-money regions.

The activation-function experiment also has a natural interpretation.
The Black-Scholes price is a smooth function of its inputs over the selected parameter ranges.
Smooth activations such as Tanh and SiLU can therefore be well suited to this regression task.
The final choice of SiLU is supported by the intermediate results and by the improved final run.

\section{Error Sources and Diagnostic Plots}

Even in a controlled synthetic setting, the neural network approximation is not exact.
Errors can arise from several sources:

\begin{itemize}
    \item finite training data;
    \item finite model capacity;
    \item imperfect optimization;
    \item numerical scaling and inverse-scaling effects;
    \item local difficulty in specific regions of the input space.
\end{itemize}
The diagnostic plots against moneyness, maturity, and volatility are therefore important.
They check whether the aggregate metrics hide localized weaknesses.
For example, if errors were systematically larger for short maturities or high volatilities, the global $R^2$ alone would not be sufficient to evaluate the model.

The project also clips negative neural-network predictions to zero after inverse scaling.
This is financially reasonable because European call prices are non-negative:

\begin{equation}
    C \geq 0.
\end{equation}

However, this correction is still a post-processing step.
It enforces a basic no-negativity constraint, but it does not guarantee all theoretical no-arbitrage properties of option prices, such as monotonicity in $S_0$ or convexity in $S_0$.

\section{Interpretation of Noisy Targets}

The noisy-target experiments should be interpreted carefully.
The noise model is:

\begin{equation}
    \tilde{C}
    =
    \max(C_{BS}+\varepsilon,0),
    \qquad
    \varepsilon
    \sim
    \mathcal{N}\left(0,\left[\ell \max(C_{BS},1)\right]^2\right).
\end{equation}

The training target is $\tilde{C}$, while evaluation remains against $C_{BS}$.
This design answers the following question:

\begin{quote}
    If the supervised labels are imperfect but still centered around the Black-Scholes price, can the model recover a good approximation of the clean pricing function?
\end{quote}

The experiments show that performance deteriorates as $\ell$ increases.
This is expected: noisy labels reduce the information available to the learner.
At the same time, moderate noise does not destroy the approximation, which suggests some robustness of the neural-network surrogate.

This experiment should not be confused with real market modeling.
Real option prices contain bid-ask spreads, liquidity effects, dividend assumptions, early-exercise considerations for some contracts, and implied-volatility structure.
The controlled noisy-target experiment is therefore a robustness analysis, not a substitute for exchange-traded option data.

\section{Computational Considerations}

The computational question in this project is not whether the pricing quantities are computable in a theoretical sense.
On the bounded synthetic domain considered in the experiments, the analytical Black-Scholes formula, the Monte Carlo estimator, the neural-network predictor, and the SVR benchmark are all numerically computable.
The relevant question is instead practical: how much computational work is required to obtain many option prices with acceptable accuracy, and how this work is distributed between training, simulation, and inference.

This distinction matters because the considered methods have different computational profiles.
Black-Scholes pricing is available in closed form, Monte Carlo pricing trades accuracy for a larger number of simulated paths, neural-network pricing pays an upfront training cost and then provides fast inference, while SVR is useful as a classical baseline but becomes expensive when the number of training samples grows.

\begin{table}[H]
    \centering
    \begin{tabular}{
        >{\raggedright\arraybackslash}p{0.27\linewidth}
        >{\raggedright\arraybackslash}p{0.33\linewidth}
        >{\raggedright\arraybackslash}p{0.30\linewidth}
        }
        \toprule
        Method         & Main computational cost       & Practical implication           \\
        \midrule
        Black-Scholes  & closed-form evaluation        & fast ground-truth computation   \\
        \projecttablerowsep
        Monte Carlo    & number of simulated paths     & accurate but expensive at scale \\
        \projecttablerowsep
        Neural network & training phase                & fast inference after training   \\
        \projecttablerowsep
        SVR            & kernel computation on samples & suitable only at reduced scale  \\
        \bottomrule
    \end{tabular}
    \caption{Cost drivers and practical implications of the pricing methods considered in the project.}
    \label{tab:computational-profile}
\end{table}

This cost perspective explains the role of the learned surrogate.
The neural network is not needed to beat the analytical Black-Scholes formula when that formula is available.
Instead, it illustrates a computational pattern that is common in surrogate modeling: an initial training cost is paid once, and subsequent evaluations are reduced to fast forward passes.
This is the perspective supported by the runtime benchmark.

\subsection{Implementation-Level Considerations}

The previous comparison also depends on how the algorithms are implemented.
As in any numerical experiment, all computations are performed with finite-precision floating-point arithmetic.
This is acceptable for the present study because the objective is not symbolic exactness, but stable numerical approximation on a controlled domain.

Python is used mainly as an orchestration language around optimized numerical libraries.
This is important for the project because the computationally expensive operations are not implemented as pure Python loops:

\begin{itemize}
    \item dataset generation and analytical pricing use vectorized NumPy operations;
    \item Monte Carlo simulation is batched;
    \item neural-network training and inference are delegated to PyTorch;
    \item the SVR benchmark relies on scikit-learn.
\end{itemize}
Therefore, the performance results reflect the behavior of a realistic scientific Python workflow based on native numerical backends, not the performance of interpreted Python code alone.

For the same reason, the Python Global Interpreter Lock is not a practical bottleneck for the current implementation.
The project does not rely on CPU-bound pure Python multithreading for its main numerical workload.
\begin{quote}
The GIL could become relevant only for future extensions that attempted to parallelize heavy computations with Python threads while keeping the work inside Python-level loops.
In that case, multiprocessing, vectorization, native extensions, or GPU-based computation would be more appropriate choices.
\end{quote}

For performance-critical components requiring finer control over memory management and parallel execution, it could also be preferable to move the computational kernel to a language better suited to low-level numerical performance, such as C++, Rust, or Julia.

\section{Limits of the Synthetic Setting}

The use of synthetic data is appropriate for the research question because the target is the Black-Scholes pricing function itself.
Using real exchange-traded option prices would change the problem.
The model would no longer learn only the mathematical map $x \mapsto C_{BS}(x)$.
Instead, it would learn a market-pricing mechanism affected by additional variables that are not present in the current input vector.

This is why the synthetic setting is not a weakness of the project.
It is a methodological choice that makes the experiment controlled, reproducible, and directly measurable.
The limitation is that the conclusions remain internal to the Black-Scholes world.

\begin{table}[H]
    \centering
    \begin{tabular}{
        >{\raggedright\arraybackslash}p{0.32\linewidth}
        >{\raggedright\arraybackslash}p{0.28\linewidth}
        >{\raggedright\arraybackslash}p{0.30\linewidth}
        }
        \toprule
        Aspect            & Synthetic setting                    & Real market setting                            \\
        \midrule
        Target            & analytical Black-Scholes price       & observed market quote                          \\
        \projecttablerowsep
        Ground truth      & known exactly                        & not directly known                             \\
        \projecttablerowsep
        Noise             & controlled by design                 & market microstructure and liquidity            \\
        \projecttablerowsep
        Evaluation        & direct error against $C_{BS}$        & depends on data quality and market assumptions \\
        \projecttablerowsep
        Research question & approximation of a known pricing map & modeling market prices                         \\
        \bottomrule
    \end{tabular}
    \caption{Difference between the project setting and a real-market option-pricing setting.}
    \label{tab:synthetic-vs-real-discussion}
\end{table}

\section{Comparison with Classical Methods}

The three main pricing approaches have different roles.

\begin{itemize}
    \item Black-Scholes is the analytical reference and ground truth.
    \item Monte Carlo is a flexible numerical estimator and runtime benchmark.
    \item The neural network is a learned pricing surrogate after training.
\end{itemize}
SVR adds a fourth perspective: a classical machine-learning baseline on reduced datasets.
It performs well, but its scalability is less favorable than the neural-network approach for larger training sets.

The final interpretation is therefore not a ranking that is valid in every situation.
It is a role-based comparison:

\begin{align}
    \text{Black-Scholes}
     & \rightarrow
    \text{ground truth},                  \\
    \text{Monte Carlo}
     & \rightarrow
    \text{classical numerical benchmark}, \\
    \text{NN}
     & \rightarrow
    \text{fast learned surrogate}.
\end{align}

This role-based interpretation keeps the conclusion precise.
The neural network is successful because it approximates the chosen pricing function accurately and evaluates quickly after training.
It is not presented as universally superior to analytical or numerical methods.
